---
title: "study 1 analysis"
output: 
  html_document:
      toc: true
      toc_float: true
      toc_depth: 3
      number_sections: true
date: "2025-12-15"
---

What this script does:
- proceeds from output of preprocessing script
- plots Sorting and Causal Task rdms, gets correlations with models RDM, gets within subject correlations 


# Setup

## load packages

```{r, setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)

library(pacman)
pacman::p_load(tidyverse,
               ggforce,
               conflicted,
               here,
               cowplot,
               lme4,
               lmerTest,
               effects,
               Hmisc,
               performance,
               purrr,
               readr,
               scales,
               patchwork,
               corrplot)

walk(c("select", "filter", "rename", "mutate", "summarise", "summarize"), ~ conflict_prefer(.x, "dplyr"))
theme_set(theme_bw())
sessionInfo()

```

## Read data and set variables

```{r}
desired_order <- c("see something", "hear something", "choose what to do", 
                   "remember something", "think about something", "reach for something", 
                   "sit down", "jump up and down", "kick something", "take a walk", 
                   "get tired", "become hungry", "feel scared", "experience pain", "get sick")

domain_order <- c("mind", "action", "bio")

freesort_columns <- c("subject_id", "x", "y", "item")
freesort_columns_with_trial_order <- c("subject_id", "x", "y", "item", "trial_type", "trial_type_order")
causation_columns <- c("subject_id", "itemA", "itemB", "response")

# paths
base_path <- here("code", "study1")
data_original_path <- paste0(base_path, "/data-combined/pilot05_121324_n50_combined_17")
snapshots_input_path <- paste0(base_path, "/snapshots_study1_preprocessing_analysis") 
snapshots_output_path <- paste0(base_path, "/snapshots_study1_final_analysis")

data_combined_path <- paste0(snapshots_input_path, "/pilot_121424_combined.csv")
theories_all_input_path <- paste0(snapshots_input_path, "/pilot_121424_theories_all.csv") 
theories_input_path <- paste0(snapshots_input_path, "/pilot_121424_theories.csv") 
correlations_output_path <- paste0(snapshots_output_path, "/correlations.csv")
rdms_input_path <- paste0(base_path, "/figures/theoretical-rdms")
rdms_outputs_path <- paste0(base_path, "/figures")

# read data
d <- read_csv(data_combined_path)

```

## Reshape data

Create separate data frames for Sorting Task, Causal Task and a combined data frame
```{r}
df_causal <- d %>%
  filter(trial_type == "causation", !attention_check, itemA != itemB) %>%
  select(all_of(causation_columns)) %>%
  mutate(
    causal_distance = (1 - response / 100 - min(1 - response / 100, na.rm = TRUE)) /
                            (max(1 - response / 100, na.rm = TRUE) - min(1 - response / 100, na.rm = TRUE))
  ) %>%
  select(subject_id, itemA, itemB, causal_distance) %>% 
  mutate(domain_itemA = case_when(itemA == "hear something" ~ "mind", #labels for first item of pair
                             itemA == "choose what to do" ~ "mind",
                             itemA == "remember something" ~ "mind",
                             itemA == "think about something" ~ "mind",
                             itemA == "see something" ~ "mind",
                             itemA == "reach for something" ~ "action",
                             itemA == "sit down" ~ "action", 
                             itemA == "jump up and down" ~ "action",
                             itemA == "kick something" ~ "action",
                             itemA == "take a walk" ~ "action",
                             itemA == "get tired" ~ "bio", 
                             itemA == "become hungry" ~ "bio",
                             itemA == "feel scared"~ "bio",
                             itemA == "experience pain"~ "bio",
                             itemA == "get sick"~ "bio"),
         domain_itemB = case_when(itemB == "hear something" ~ "mind", 
                               itemB == "choose what to do" ~ "mind",
                               itemB == "remember something" ~ "mind",
                               itemB == "think about something" ~ "mind",
                               itemB == "see something" ~ "mind",
                               itemB == "reach for something" ~ "action",
                               itemB == "sit down" ~ "action", 
                               itemB == "jump up and down" ~ "action",
                               itemB == "kick something" ~ "action",
                               itemB == "take a walk" ~ "action",
                               itemB == "get tired" ~ "bio", 
                               itemB == "become hungry" ~ "bio",
                               itemB == "feel scared"~ "bio",
                               itemB == "experience pain"~ "bio",
                               itemB == "get sick"~ "bio")) %>% 
  relocate(domain_itemA, .after = "itemB") %>% 
  relocate(domain_itemB, .after = "domain_itemA") %>% 
  mutate(itemA = factor(itemA, levels = desired_order),
         itemB = factor(itemB, levels = desired_order),
         domain_itemA = factor(domain_itemA, levels = domain_order),
         domain_itemB = factor(domain_itemB, levels = domain_order))


df_freesort <- d %>%
  filter(trial_type == "free_sort") %>%
  select(all_of(freesort_columns_with_trial_order)) %>%
  rename(item_freesort = item) %>%
  
  # Expand pairs within each subject
  group_by(subject_id, trial_type_order) %>%
  summarize(pairs = list(expand.grid(itemA = item_freesort, itemB = item_freesort, stringsAsFactors = FALSE)), .groups = "drop") %>%
  unnest(pairs) %>%
  filter(itemA != itemB) %>%
  
  # Join Free Sort coordinates
  left_join(d %>% filter(trial_type == "free_sort") %>%
              select(subject_id, itemA = item, x, y),
            by = c("subject_id", "itemA")) %>%
  rename(x_itemA = x, y_itemA = y) %>%
  
  left_join(d %>% filter(trial_type == "free_sort") %>%
              select(subject_id, itemB = item, x, y),
            by = c("subject_id", "itemB")) %>%
  rename(x_itemB = x, y_itemB = y) %>%
  
  # Compute Euclidean Distance and Normalize
  mutate(
    euclidean_distance = sqrt((x_itemA - x_itemB)^2 + (y_itemA - y_itemB)^2),
    freesort_distance = round( (euclidean_distance - min(euclidean_distance, na.rm = TRUE)) /
                    (max(euclidean_distance, na.rm = TRUE) - min(euclidean_distance, na.rm = TRUE)),2),
    
    trial_type_order = ifelse(trial_type_order == "first", "causation_second", "causation_first"),
    
    itemA = factor(itemA, levels = desired_order),
    itemB = factor(itemB, levels = desired_order)
  ) %>%
  select(subject_id, itemA, itemB, freesort_distance, trial_type_order) %>%
  ungroup()

df_combined <- df_freesort %>%
  left_join(df_causal, by = c("subject_id", "itemA", "itemB")) %>% 
  relocate(freesort_distance, .before = "causal_distance") %>% 
  relocate(trial_type_order, .after = "domain_itemB")

write.csv(df_combined, paste0(snapshots_output_path, "/study1_processed.csv"), row.names = F)
```

# Freesort RDMs

## plot subject-specific raw responses

```{r, fig.width = 14, fig.height = 10}
d %>% 
  filter(trial_type == "free_sort") %>%
  mutate(y = 500 - y) %>% #invert y axis
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_text(aes(label = item), size = 3, vjust = -.2) +
  facet_wrap(~subject_id)

```

## plot average raw responses

```{r}
# pdf("my_plot_average.pdf", width = 7, height = 4)  # Width and height are in inches
d %>% 
  filter(trial_type == "free_sort") %>%
  mutate(y = 500 - y) %>% 
  group_by(item) %>% 
  mutate(mean_x = mean(x),
            mean_y = mean(y)) %>% 
  ungroup() %>% 
  distinct(item, .keep_all = TRUE) %>% 
  select(-subject_id) %>% 
  ggplot(aes(x = mean_x, y = mean_y)) +
  geom_point() +
  geom_text(aes(label = item), vjust = -1) +
  labs(title = "Average Raw Responses")

```

## plot subject-specific Sorting Task rdms

```{r}
plot_rdm <- function(data, subject, desired_order) {
  df_matrix <- data %>%
    filter(subject_id == subject) %>%
    select(-subject_id) %>%
    select(itemA, itemB, freesort_distance) %>% 
    pivot_wider(names_from = itemB, values_from = freesort_distance) %>%
    column_to_rownames("itemA")
  
  df_matrix <- df_matrix[desired_order, desired_order]
  diag(df_matrix) <- 0 
  
  corrplot(as.matrix(df_matrix), method = "color", is.corr = FALSE,
    tl.col = "black", title = paste("RDM - ", subject),
    mar = c(0, 0, 2, 0), addgrid.col = "darkgray"
  )
}

subject_ids <- unique(df_freesort$subject_id)

#generate plots
# plots <- map(subject_ids, ~plot_rdm(df_freesort, .x, desired_order)) #run this to generate plots

#save plots with subject name as file name
new_names <- paste0("subj_", str_pad(seq_along(subject_ids), width = 2, pad = "0")) 

#loop to save each plot as a separate PDF
for (i in seq_along(subject_ids)) {  
  pdf(file = paste0(base_path, "/figures/freesort/", new_names[i], ".pdf"), width = 8, height = 6)  #open a PDF device
  plot_rdm(df_freesort, subject_ids[i], desired_order)  
  dev.off()  #close the device to save the file
}

```

## figure for paper: plot average freesort rdm 

```{r, fig.width = 12, fig.height = 12}
df_mean_freesort <- df_freesort %>% 
  unite("item_pairs", itemA, itemB, sep = "_", remove = FALSE) %>% 
  group_by(item_pairs) %>% 
  mutate(mean_freesort_distance = mean(freesort_distance)) %>% 
  distinct(item_pairs, .keep_all = TRUE) %>% 
  ungroup() %>% 
  select(subject_id, itemA, itemB, mean_freesort_distance)  %>% 
  mutate(itemA = factor(itemA, levels = desired_order),
         itemB = factor(itemB, levels = desired_order))
  
# matrixify and plot
df_mean_freesort_rdm <- df_mean_freesort %>% 
  pivot_wider(names_from = itemB, values_from = mean_freesort_distance) %>% 
  column_to_rownames("itemA")  
  
df_mean_freesort_matrix <- df_mean_freesort_rdm[desired_order, desired_order]

diag(df_mean_freesort_matrix) <- NA #set diagonals 

#min max normalize
min_val <- min(df_mean_freesort_matrix, na.rm = TRUE) 
max_val <- max(df_mean_freesort_matrix, na.rm = TRUE) 
df_mean_freesort_matrix <- (df_mean_freesort_matrix - min_val) / (max_val - min_val) 

group_colors <- c("#2F2585", "#69ad44", "#f36b2d") 

#map each item to a color based on its group
item_colors <- setNames(c(rep(group_colors[1], 5), 
                          rep(group_colors[2], 5), 
                          rep(group_colors[3], 5)), 
                        desired_order)

#ensure colors align with the matrix order
label_colors <- item_colors[rownames(df_mean_freesort_matrix)]

custom_palette <- colorRampPalette(c("red", "white", "blue")) 

pdf(file = paste0(rdms_outputs_path, "/freesort/averageEuclideanRDM.pdf"), width = 12, height = 12)

#large image
corrplot(
  as.matrix(df_mean_freesort_matrix),
  method = "color",
  is.corr = FALSE,
  col = custom_palette(200),
  #tl.col = "black", #could be label_colors
  tl.col = label_colors, #could be label_colors
  tl.cex = 2,
  na.label = "square",       
  na.label.col = "darkgrey",   
  #title = "Average Freesort RDM",
  mar = c(0, 0, 2, 0),
  addgrid.col = "darkgray",
  cl.offset = 1.0,  # Push labels even further
  cl.cex = 2,     # Slightly smaller label text
  cl.ratio = 0.22,   # Widen the legend bar. Change this to push words far away
  cex.main = 3
)

# legend for the domains
legend(
  "topleft",
  inset = c(0.1, 0.1), #move the legend outside the plot
  legend = c("Mind", "Action", "Body"), 
  fill = group_colors, 
  border = NA, 
  bty = "n", #no box around the legend
  cex = 2, #text size
  xpd = TRUE, #allow drawing outside the plot area
  #ncol = 1,
  y.intersp = 1.4 #increase this for more space between rows
)

```

# Causal RDMs

## plot subject-specific causal RDMs

```{r}
plot_causal_rdm <- function(data, subject, desired_order) {
  
  df_matrix <- data %>%
    filter(subject_id == subject) %>%
    select(-subject_id) %>%
    select(itemA, itemB, causal_distance) %>% 
    pivot_wider(names_from = itemB, values_from = causal_distance) %>%
    column_to_rownames("itemA")
  
  df_matrix <- df_matrix[desired_order, desired_order]
  diag(df_matrix) <- 0
  
  corrplot(as.matrix(df_matrix), method = "color", is.corr = FALSE,
           tl.col = "black", title = paste("Causal RDM - ", subject),
           mar = c(0, 0, 2, 0), addgrid.col = "darkgray")
}

subject_ids <- unique(df_causal$subject_id)

#generate plots
# plots <- map(subject_ids, ~plot_causal_rdm(df_causal, .x, desired_order)) #run this to generate plots

# save plots with subject name as file name
new_names <- paste0("subj_", str_pad(seq_along(subject_ids), width = 2, pad = "0"))

for (i in seq_along(subject_ids)) { 

    pdf(file = paste0(base_path, "/figures/causal/", new_names[i], ".pdf"), width = 8, height = 6)
  plot_causal_rdm(df_causal, subject_ids[i], desired_order)  
  dev.off()  
}

```

## figure for paper: plot average causal RDM

the tris show a mixture of what you are causing and what is being caused. the more top right you are, the more you're showing what you cause, and the lower you are, the more you're showing being caused
top half of items: your the cause mostly, middle item: equal likelihood, bottom half: your ethe effect mostly

```{r, fig.width = 12, fig.height = 12}
df_mean_causal <- df_causal %>% 
  mutate(item_pairs = str_c(itemA, itemB, sep = "-"))  %>%
  group_by(item_pairs) %>%
  mutate(mean_causal_distance = mean(causal_distance, na.rm = TRUE)) %>%  #average across participants
  ungroup() 

df_mean_causal_rdm <- df_mean_causal %>%
  distinct(item_pairs, .keep_all = TRUE) %>% 
  select(itemA, itemB, mean_causal_distance) %>% 
  pivot_wider(names_from = itemB, values_from = mean_causal_distance) %>%
  column_to_rownames("itemA")  

df_mean_causal_rdm <- df_mean_causal_rdm[desired_order, desired_order]

diag(df_mean_causal_rdm) <- NA 

#min max normalize
df_mean_causal_matrix <- as.matrix(df_mean_causal_rdm) 
min_val <- min(df_mean_causal_matrix, na.rm = TRUE) 
max_val <- max(df_mean_causal_matrix, na.rm = TRUE) 
df_mean_causal_matrix <- (df_mean_causal_matrix - min_val) / (max_val - min_val) 

#colors
group_colors <- c("#f7bf2a", "#69ad44", "#f36b2d") 

item_colors <- setNames(c(rep(group_colors[1], 5), #mind
                          rep(group_colors[2], 5), #action
                          rep(group_colors[3], 5)), #body
                        desired_order)

label_colors <- item_colors[rownames(df_mean_causal_rdm)]

custom_palette <- colorRampPalette(c("red", "white", "blue")) 

pdf(file = paste0(rdms_outputs_path, "/causal/averageCausalRDM.pdf"), width = 12, height = 12)

#large image
corrplot(
  df_mean_causal_matrix,
  method = "color",
  is.corr = FALSE,
  col = custom_palette(200),
  #tl.col = "black", 
  tl.col = label_colors, 
  tl.cex = 2,
  mar = c(0, 0, 2, 0),
  addgrid.col = "darkgray",
  na.label = "square",       
  na.label.col = "darkgrey",
  cl.offset = 1.0, 
  cl.cex = 2, #label text
  cl.ratio = 0.22, #push words far away
  cex.main = 3
)

legend(
  "topleft", 
  inset = c(0.1, 0.1), 
  legend = c("Mind", "Action", "Body"),
  fill = group_colors, 
  border = NA, 
  bty = "n", 
  cex = 2, #text size
  xpd = TRUE, #draw outside the plot area
  #ncol = 1,
  y.intersp = 1.4 #more space between rows
)

```




## Specific causal relations for paper 

```{r}
#mind->action
df_filtered <- df_causal %>% 
  filter(domain_itemA == "mind" & domain_itemB == "action")
mean_distance <- mean(df_filtered$causal_distance, na.rm = TRUE)
sem <- sd(df_filtered$causal_distance, na.rm = TRUE) / sqrt(nrow(df_filtered))
ci_lower <- mean_distance - qt(0.975, df = nrow(df_filtered) - 1) * sem
ci_upper <- mean_distance + qt(0.975, df = nrow(df_filtered) - 1) * sem
mean_distance
ci_lower
ci_upper

#action->mind
df_filtered <- df_causal %>% 
  filter(domain_itemA == "action" & domain_itemB == "mind")
mean_distance <- mean(df_filtered$causal_distance, na.rm = TRUE)
sem <- sd(df_filtered$causal_distance, na.rm = TRUE) / sqrt(nrow(df_filtered))
ci_lower <- mean_distance - qt(0.975, df = nrow(df_filtered) - 1) * sem
ci_upper <- mean_distance + qt(0.975, df = nrow(df_filtered) - 1) * sem
mean_distance
ci_lower
ci_upper

#action->action
df_filtered <- df_causal %>% 
  filter(domain_itemA == "action" & domain_itemB == "action")
mean_distance <- mean(df_filtered$causal_distance, na.rm = TRUE)
sem <- sd(df_filtered$causal_distance, na.rm = TRUE) / sqrt(nrow(df_filtered))
ci_lower <- mean_distance - qt(0.975, df = nrow(df_filtered) - 1) * sem
ci_upper <- mean_distance + qt(0.975, df = nrow(df_filtered) - 1) * sem
mean_distance
ci_lower
ci_upper

#body->body
df_filtered <- df_causal %>% 
  filter(domain_itemA == "bio" & domain_itemB == "bio")
mean_distance <- mean(df_filtered$causal_distance, na.rm = TRUE)
sem <- sd(df_filtered$causal_distance, na.rm = TRUE) / sqrt(nrow(df_filtered))
ci_lower <- mean_distance - qt(0.975, df = nrow(df_filtered) - 1) * sem
ci_upper <- mean_distance + qt(0.975, df = nrow(df_filtered) - 1) * sem
mean_distance
ci_lower
ci_upper

#body->actions
df_filtered <- df_causal %>% 
  filter(domain_itemA == "bio" & domain_itemB == "action")
mean_distance <- mean(df_filtered$causal_distance, na.rm = TRUE)
sem <- sd(df_filtered$causal_distance, na.rm = TRUE) / sqrt(nrow(df_filtered))
ci_lower <- mean_distance - qt(0.975, df = nrow(df_filtered) - 1) * sem
ci_upper <- mean_distance + qt(0.975, df = nrow(df_filtered) - 1) * sem
mean_distance
ci_lower
ci_upper

#actions->body
df_filtered <- df_causal %>% 
  filter(domain_itemA == "action" & domain_itemB == "bio")
mean_distance <- mean(df_filtered$causal_distance, na.rm = TRUE)
sem <- sd(df_filtered$causal_distance, na.rm = TRUE) / sqrt(nrow(df_filtered))
ci_lower <- mean_distance - qt(0.975, df = nrow(df_filtered) - 1) * sem
ci_upper <- mean_distance + qt(0.975, df = nrow(df_filtered) - 1) * sem
mean_distance
ci_lower
ci_upper

#perception->cognition
df_filtered <- df_causal %>% 
  filter(domain_itemA == "mind" & domain_itemB == "mind") %>% 
  filter(itemA %in% c("hear something", "see something") &
         itemB %in% c("choose what to do", "think about something", "remember something"))
mean_distance <- mean(df_filtered$causal_distance, na.rm = TRUE)
sem <- sd(df_filtered$causal_distance, na.rm = TRUE) / sqrt(nrow(df_filtered))
ci_lower <- mean_distance - qt(0.975, df = nrow(df_filtered) - 1) * sem
ci_upper <- mean_distance + qt(0.975, df = nrow(df_filtered) - 1) * sem
mean_distance
ci_lower
ci_upper

#cognition->perception
df_filtered <- df_causal %>% 
  filter(domain_itemA == "mind" & domain_itemB == "mind") %>% 
  filter(itemB %in% c("hear something", "see something") &
         itemA %in% c("choose what to do", "think about something", "remember something"))
mean_distance <- mean(df_filtered$causal_distance, na.rm = TRUE)
sem <- sd(df_filtered$causal_distance, na.rm = TRUE) / sqrt(nrow(df_filtered))
ci_lower <- mean_distance - qt(0.975, df = nrow(df_filtered) - 1) * sem
ci_upper <- mean_distance + qt(0.975, df = nrow(df_filtered) - 1) * sem
mean_distance
ci_lower
ci_upper

#seeing->sick
df_filtered <- df_causal %>% 
  filter(itemA == "see something" & itemB == "get sick")
mean_distance <- mean(df_filtered$causal_distance, na.rm = TRUE)
sem <- sd(df_filtered$causal_distance, na.rm = TRUE) / sqrt(nrow(df_filtered))
ci_lower <- mean_distance - qt(0.975, df = nrow(df_filtered) - 1) * sem
ci_upper <- mean_distance + qt(0.975, df = nrow(df_filtered) - 1) * sem
mean_distance
ci_lower
ci_upper

#hear->sick
df_filtered <- df_causal %>% 
  filter(itemA == "hear something" & itemB == "get sick")
mean_distance <- mean(df_filtered$causal_distance, na.rm = TRUE)
sem <- sd(df_filtered$causal_distance, na.rm = TRUE) / sqrt(nrow(df_filtered))
ci_lower <- mean_distance - qt(0.975, df = nrow(df_filtered) - 1) * sem
ci_upper <- mean_distance + qt(0.975, df = nrow(df_filtered) - 1) * sem
mean_distance
ci_lower
ci_upper

```


# Correlations: Sorting vs Causal Task

## mean within-subject correlation

```{r}
compute_subjectwise_rdm_correlation <- function(df_causal, df_freesort, desired_order, metric = "kendall") {
  subject_ids <- unique(df_causal$subject_id)
  subject_correlations <- numeric(length(subject_ids))
  
  for (i in seq_along(subject_ids)) {
    subject <- subject_ids[i]
    
    freesort_rdm <- df_freesort %>%
      filter(subject_id == subject) %>%
      select(itemA, itemB, freesort_distance) %>%
      pivot_wider(names_from = itemB, values_from = freesort_distance) %>%
      column_to_rownames("itemA") %>%
      as.matrix()
    freesort_rdm <- freesort_rdm[desired_order, desired_order]
    
    causal_rdm <- df_causal %>%
      filter(subject_id == subject) %>%
      select(itemA, itemB, causal_distance) %>%
      pivot_wider(names_from = itemB, values_from = causal_distance) %>%
      column_to_rownames("itemA") %>%
      as.matrix()
    causal_rdm <- causal_rdm[desired_order, desired_order]
    
    freesort_flat <- as.numeric(freesort_rdm)
    causal_flat <- as.numeric(causal_rdm)
    
    subject_correlations[i] <- cor(freesort_flat, causal_flat, method = metric, use = "complete.obs")
  }
  
  return(subject_correlations)
}

subject_correlations <- compute_subjectwise_rdm_correlation(df_causal, df_freesort, desired_order)
mean_correlation <- mean(subject_correlations, na.rm = TRUE)
mean_correlation
```

## t-test

```{r}
t_test_result <- t.test(subject_correlations, mu = 0)
t_test_result
```

## t-test assumptions

seem fine, so no need for perm test
```{r}
shapiro_result <- shapiro.test(subject_correlations)
print(shapiro_result)

qqnorm(subject_correlations)
qqline(subject_correlations, col = "red")

hist(subject_correlations, main = "Histogram of Subject Correlations", 
     xlab = "Correlation", col = "lightblue", border = "black")
```

# Correlations: Data vs Models

## General functions for correlation plot

```{r}
extract_upper_tri <- function(matrix) {
  matrix[upper.tri(matrix)]
}

compute_rdm_correlation <- function(data, emp_col, theor_col, desired_order, method = "kendall") {
  rdm1 <- data %>%
    select(itemA, itemB, {{ emp_col }}) %>%
    pivot_wider(names_from = itemB, values_from = {{ emp_col }}) %>%
    column_to_rownames("itemA") %>%
    as.matrix()
  
  rdm2 <- data %>%
    select(itemA, itemB, {{ theor_col }}) %>%
    pivot_wider(names_from = itemB, values_from = {{ theor_col }}) %>%
    column_to_rownames("itemA") %>%
    as.matrix()
  
  rdm1 <- rdm1[desired_order, desired_order]
  rdm2 <- rdm2[desired_order, desired_order]
  
  diag(rdm1) <- NA
  diag(rdm2) <- NA
  
  upper_rdm1 <- extract_upper_tri(rdm1)
  upper_rdm2 <- extract_upper_tri(rdm2)
  
  cor(upper_rdm1, upper_rdm2, method = method, use = "complete.obs")
}

```

## noise ceiling functions

```{r}
calculate_freesort_noise_ceiling <- function(data, subjects, desired_order, metric = "kendall") {
  # Initialize storage for correlations
  upper_bound <- numeric(length(subjects))
  lower_bound <- numeric(length(subjects))
  
  for (i in seq_along(subjects)) {
    subject <- subjects[i]
    
    #Subject's RDM
    subject_rdm <- data %>%
        filter(subject_id == subject) %>%
        select(subject_id, itemA, itemB, freesort_distance) %>% 
        pivot_wider(names_from = itemB, values_from = freesort_distance) %>%
        column_to_rownames("itemA") %>%
        as.matrix()
    subject_rdm <- subject_rdm[desired_order, desired_order]
    
    #Grand average RDM
    grand_avg_rdm <- data %>%
        select(subject_id, itemA, itemB, freesort_distance) %>% 
        unite("item_pairs", itemA, itemB, sep = "_", remove = FALSE) %>% 
        group_by(item_pairs) %>%
        mutate(freesort_distance = mean(freesort_distance)) %>%
        ungroup() %>% 
        distinct(item_pairs, .keep_all = TRUE) %>% 
        select(-c(item_pairs, subject_id)) %>% 
        pivot_wider(names_from = itemB, values_from = freesort_distance) %>%
        column_to_rownames("itemA") %>%
        as.matrix()
    grand_avg_rdm <- grand_avg_rdm[desired_order, desired_order]
    
    #Other subjects' average RDM
    other_avg_rdm <- data %>%
      select(subject_id, itemA, itemB, freesort_distance) %>% 
      filter(subject_id != "subj_01") %>%
      unite("item_pairs", itemA, itemB, sep = "_", remove = FALSE) %>% 
      group_by(item_pairs) %>%
      mutate(freesort_distance = mean(freesort_distance)) %>%
      ungroup() %>% 
      distinct(item_pairs, .keep_all = TRUE) %>% 
      select(-c(item_pairs, subject_id)) %>% 
      pivot_wider(names_from = itemB, values_from = freesort_distance) %>%
      column_to_rownames("itemA") %>%
      as.matrix()
    other_avg_rdm <- other_avg_rdm[desired_order, desired_order]
    
    subject_upper <- as.numeric(subject_rdm[upper.tri(subject_rdm)])
    other_upper <- as.numeric(other_avg_rdm[upper.tri(other_avg_rdm)])
    grand_upper <- as.numeric(grand_avg_rdm[upper.tri(grand_avg_rdm)])
    
    upper_bound[i] <- cor(subject_upper, other_upper, method = metric)
    lower_bound[i] <- cor(subject_upper, grand_upper, method = metric)
  }
  
  # Return average bounds
  list(
    upper_bound = mean(upper_bound, na.rm = TRUE),
    lower_bound = mean(lower_bound, na.rm = TRUE)
  )
}

calculate_causal_noise_ceiling <- function(data, subjects, desired_order, metric = "kendall") {
  #Initialize storage for correlations
  upper_tri_upper_bound <- numeric(length(subjects))
  upper_tri_lower_bound <- numeric(length(subjects))
  
  lower_tri_upper_bound <- numeric(length(subjects))
  lower_tri_lower_bound <- numeric(length(subjects))
  
  #Loop through each subject
  for (i in seq_along(subjects)) {
    subject <- subjects[i]
    
    #subject's RDM
    subject_rdm <- data %>%
        filter(subject_id == subject) %>%
        select(subject_id, itemA, itemB, causal_distance) %>% 
        pivot_wider(names_from = itemB, values_from = causal_distance) %>%
        column_to_rownames("itemA") %>%
        as.matrix()
    subject_rdm <- subject_rdm[desired_order, desired_order]
    
    #grand average RDM
    grand_avg_rdm <- data %>%
        select(subject_id, itemA, itemB, causal_distance) %>% 
        unite("item_pairs", itemA, itemB, sep = "_", remove = FALSE) %>% 
        group_by(item_pairs) %>%
        mutate(causal_distance = mean(causal_distance)) %>%
        ungroup() %>% 
        distinct(item_pairs, .keep_all = TRUE) %>% 
        select(-c(item_pairs, subject_id)) %>% 
        pivot_wider(names_from = itemB, values_from = causal_distance) %>%
        column_to_rownames("itemA") %>%
        as.matrix()
    grand_avg_rdm <- grand_avg_rdm[desired_order, desired_order]
    
    #other subjects' average RDM
    other_avg_rdm <- data %>%
      select(subject_id, itemA, itemB, causal_distance) %>% 
      filter(subject_id != "subj_01") %>%
      unite("item_pairs", itemA, itemB, sep = "_", remove = FALSE) %>% 
      group_by(item_pairs) %>%
      mutate(causal_distance = mean(causal_distance)) %>%
      ungroup() %>% 
      distinct(item_pairs, .keep_all = TRUE) %>% 
      select(-c(item_pairs, subject_id)) %>% 
      pivot_wider(names_from = itemB, values_from = causal_distance) %>%
      column_to_rownames("itemA") %>%
      as.matrix()
    other_avg_rdm <- other_avg_rdm[desired_order, desired_order]
    
    #upper tri noise ceiling
    subject_upper <- as.numeric(subject_rdm[upper.tri(subject_rdm)])
    other_upper <- as.numeric(other_avg_rdm[upper.tri(other_avg_rdm)])
    grand_upper <- as.numeric(grand_avg_rdm[upper.tri(grand_avg_rdm)])

    upper_tri_upper_bound[i] <- cor(subject_upper, other_upper, method = metric)
    upper_tri_lower_bound[i] <- cor(subject_upper, grand_upper, method = metric)

    #lower tri noise ceiling
    subject_lower <- as.numeric(subject_rdm[lower.tri(subject_rdm)])
    other_lower <- as.numeric(other_avg_rdm[lower.tri(other_avg_rdm)])
    grand_lower <- as.numeric(grand_avg_rdm[lower.tri(grand_avg_rdm)])
    
    lower_tri_upper_bound[i] <- cor(subject_lower, other_lower, method = metric)
    lower_tri_lower_bound[i] <- cor(subject_lower, grand_lower, method = metric)

  }
  
  #return average bounds
  list(
    upper_tri_upper_bound = mean(upper_tri_upper_bound, na.rm = TRUE),
    upper_tri_lower_bound = mean(upper_tri_lower_bound, na.rm = TRUE),
    
    lower_tri_upper_bound = mean(lower_tri_upper_bound, na.rm = TRUE),
    lower_tri_lower_bound = mean(lower_tri_lower_bound, na.rm = TRUE)
  )
}

```

## Calculate Sorting Task noise ceiling

```{r}
#create correlations data frame
d_all_theories <- read_csv(theories_all_input_path)

d_freesort_correlations <- df_combined %>%
  left_join(d_all_theories, by = c("itemA", "itemB")) %>%
  group_by(subject_id) %>%
  mutate(
    `1_Mind-Body-Action` = compute_rdm_correlation(pick(everything()), freesort_distance, `theor1_Mind-Body-Action`, desired_order),
    `2_PerCog-ObjDir-Stimulus` = compute_rdm_correlation(pick(everything()), freesort_distance, `theor2_PerCog-OdActSpAct-BodyStimBodynoStim`, desired_order),
    `3_Physical-Ethereal` = compute_rdm_correlation(pick(everything()), freesort_distance, `theor3_Physical-Ethereal`, desired_order),
    `4_cosine_similarity` = compute_rdm_correlation(pick(everything()), freesort_distance, `theor4_cosine_similarity`, desired_order),
    `5_causal_judgments` = compute_rdm_correlation(pick(everything()), freesort_distance, `causal_distance`, desired_order)
  ) %>%
  ungroup() %>%
  relocate(causal_distance, .after = "freesort_distance")

subjects_freesort <- unique(d_freesort_correlations$subject_id)
noise_ceiling_sorting <- calculate_freesort_noise_ceiling(d_freesort_correlations, subjects_freesort, desired_order)
noise_ceiling_sorting
```

## Calculate Causal Task noise ceiling

```{r}
d_causal_correlations <-  df_combined %>%
  left_join(d_all_theories, by = c("itemA", "itemB")) %>% 
  group_by(subject_id) %>%
  mutate(
    `1_Mind-Body-Action` = compute_rdm_correlation(pick(everything()), causal_distance, `theor1_Mind-Body-Action`, desired_order),
    `2_PerCog-ObjDir-Stimulus` = compute_rdm_correlation(pick(everything()), causal_distance, `theor2_PerCog-OdActSpAct-BodyStimBodynoStim`, desired_order),
    `3_Physical-Ethereal` = compute_rdm_correlation(pick(everything()), causal_distance, `theor3_Physical-Ethereal`, desired_order),
    `4_cosine_similarity` = compute_rdm_correlation(pick(everything()), causal_distance, `theor4_cosine_similarity`, desired_order),
    `5_causal_judgments` = compute_rdm_correlation(pick(everything()), causal_distance, `causal_distance`, desired_order)
  ) %>%
  ungroup() 

# Compute noise ceiling for all subjects
subjects_causal <- unique(d_causal_correlations$subject_id)
noise_ceiling_causal <- calculate_causal_noise_ceiling(d_causal_correlations,subjects_causal, desired_order)
noise_ceiling_causal

```

## figure: plot correlations both tasks

```{r, fig.height = 10, fig.width = 12}
aggregated_freesort_correlations <- d_freesort_correlations %>% 
  rename("3 Category" = `1_Mind-Body-Action`,
         "6 Category" = `2_PerCog-ObjDir-Stimulus`,
         "2 Category" = `3_Physical-Ethereal`,
         "Cosine Similarity" = `4_cosine_similarity`) %>% 
  pivot_longer(
    cols = c("2 Category", "3 Category", "6 Category", "Cosine Similarity"),
    names_to = "correlation_type",
    values_to = "correlation_value"
  ) %>%
  group_by(subject_id, correlation_type) %>%
  summarize(mean_correlation = mean(correlation_value, na.rm = TRUE), .groups = "drop") %>% 
  mutate(method = "Sorting Task")

aggregated_causal_correlations <- d_causal_correlations %>% 
  rename("3 Category" = `1_Mind-Body-Action`,
         "6 Category" = `2_PerCog-ObjDir-Stimulus`,
         "2 Category" = `3_Physical-Ethereal`,
         "Cosine Similarity" = `4_cosine_similarity`) %>% 
  pivot_longer(
    cols = c("2 Category", "3 Category", "6 Category", "Cosine Similarity"),
    names_to = "correlation_type",
    values_to = "correlation_value"
  ) %>%
  group_by(subject_id, correlation_type) %>%
  summarize(mean_correlation = mean(correlation_value, na.rm = TRUE), .groups = "drop") %>% 
  mutate(method = "Causal Task")

#combine both corrs into one
d_combined_correlations <- bind_rows(aggregated_freesort_correlations, aggregated_causal_correlations) %>% 
  mutate(method = factor(method, levels = c("Sorting Task", "Causal Task"))) 

bar_plot <- ggplot(d_combined_correlations, 
                   aes(x = reorder(correlation_type, mean_correlation, FUN = function(x) -mean(x)), 
                        y = mean_correlation, fill = method)) +
  stat_summary(
    fun = "mean",
    geom = "bar",
    position = "dodge",  
    color = "black"
  ) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.3,
    position = position_dodge(width = 0.9), 
    color = "black"
  ) +
  geom_point(aes(color = method), alpha = 0.3, size = 3, position = position_jitterdodge(jitter.width = 0.2)) +
  scale_fill_manual(values = c("Sorting Task" = "darkgray", "Causal Task" = "black")) + #bar fill
  scale_color_manual(values = c("Sorting Task" = "lightgray", "Causal Task" = "#4A4A4A")) + #point colors
  
  geom_hline(yintercept = noise_ceiling_sorting$upper_bound, color = "darkgray", linetype = "dashed", linewidth = 0.5) +
  geom_hline(yintercept = noise_ceiling_sorting$lower_bound, color = "darkgray", linetype = "dashed", linewidth = 0.5) +
  geom_hline(yintercept = noise_ceiling_causal$upper_bound, color = "black", linetype = "dashed", linewidth = 0.5) +
  geom_hline(yintercept = noise_ceiling_causal$lower_bound, color = "black", linetype = "dashed", linewidth = 0.5) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 16),
    axis.ticks.length = unit(.25, "cm"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 18),
    legend.position = "top",
    legend.title = element_blank(),
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5)  
  ) +
  theme(
    plot.margin = margin(t = 10, r = 10, b = 20, l = 20),
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5)  # Adjust main title size
  ) +
  labs(
    y = "Mean RDM Correlation",
    title = "Comparison of Theoretical and Empirical RDM Correlations"
  )

# Load and create RDM images
rdm_2_category <- ggdraw() + draw_image(paste0(rdms_input_path, "/theory_4.pdf"), scale = 1.2)
rdm_3_category <- ggdraw() + draw_image(paste0(rdms_input_path, "/theory_2.pdf"), scale = 1.2)
rdm_6_category <- ggdraw() + draw_image(paste0(rdms_input_path, "/theory_3.pdf"), scale = 1.2)
rdm_cosine <- ggdraw() + draw_image(paste0(rdms_input_path, "/theory_1.pdf"), scale = 1.2)

rdm_row <- rdm_3_category + rdm_6_category + rdm_2_category + rdm_cosine +
  plot_layout(ncol = 4)

plot_correlations <- bar_plot / rdm_row + 
  plot_layout(heights = c(5, 2)) #adjusts height ratio

ggsave(
  filename = "figures/rdm_correlation_plot.pdf",
  plot = plot_correlations,
  device = "pdf",
  #path = "path/to/directory",
  width = 12,
  height = 5,
  units = "in"
)

```

## t-test

```{r}
t_test_results <- d_combined_correlations %>%
  group_by(method, correlation_type) %>%
  dplyr::summarise(
    mean_mean_correlation = mean(mean_correlation, na.rm = TRUE),
    t_value = t.test(mean_correlation, mu = 0)$statistic,  
    p_value = t.test(mean_correlation, mu = 0)$p.value,
    .groups = "drop"
  )

print(t_test_results)

```

# Item-selection for study 2

```{r}
#get mean distances by item pairs
df_pair_means <- df_combined %>% 
  select(subject_id, itemA, itemB, freesort_distance, causal_distance) %>% 
  unite("item_pairs", itemA, itemB, sep = "_", remove = FALSE) %>% 
  group_by(item_pairs) %>%
  summarize(mean_causal_distance = mean(causal_distance, na.rm = TRUE),
         mean_freesort_distance = mean(freesort_distance, na.rm = TRUE)) %>% 
  ungroup() %>% 
  separate(item_pairs, c("itemA", "itemB"), sep = "_")

#get itemB pair means
df_itemB_means <- df_pair_means %>% 
  group_by(itemB) %>%
  mutate(mean_itemB_causal_distance = mean(mean_causal_distance, na.rm = TRUE),
         mean_itemB_freesort_distance = mean(mean_freesort_distance, na.rm = TRUE)) %>%
  ungroup() 

#add domain labels
df_itemB_means_domain <- df_itemB_means %>% 
  left_join(df_combined %>% 
              select(itemA, itemB, domain_itemA, domain_itemB) %>% 
              distinct(),
              by = c("itemA", "itemB")) %>% 
  relocate(domain_itemA, .after = "itemB") %>% 
  relocate(domain_itemB, .after = "domain_itemA") %>% 
  mutate(itemB = factor(itemB, levels = desired_order))

df_itemB_means_domain_long <- df_itemB_means_domain %>% 
  pivot_longer(
    cols = c("mean_causal_distance", "mean_freesort_distance"),
    names_to = "mean_type",
    values_to = "values"
  ) %>% 
  select(-c(mean_itemB_causal_distance, mean_itemB_freesort_distance))

df_distance_differences <- df_itemB_means_domain_long %>% 
  pivot_wider(names_from = "mean_type",
              values_from = "values") %>% 
  mutate(freesort_minus_causal = mean_freesort_distance - mean_causal_distance)

write.csv(df_distance_differences, paste0(snapshots_output_path, "/relative_distances.csv"), row.names = F)
```

# Order effects

## causal-causal vs freesort-freesort correlation

```{r}
df_causal_first <- df_combined %>% 
  filter(trial_type_order == "causation_first")

df_freesort_first <- df_combined %>% 
  filter(trial_type_order == "causation_second")

# Getting causal RDMs
## causation-first group
df_causal_rdm_causation_first <- df_causal_first %>%
  mutate(item_pairs = str_c(itemA, itemB, sep = "-")) %>%
  group_by(item_pairs) %>%
  mutate(mean_causal = mean(causal_distance, na.rm = TRUE)) %>%
  ungroup() %>%
  distinct(item_pairs, .keep_all = TRUE) %>%
  select(itemA, itemB, mean_causal) %>%
  pivot_wider(names_from = itemB, values_from = mean_causal) %>%
  column_to_rownames("itemA")

df_causal_rdm_causation_first <- df_causal_rdm_causation_first[desired_order, desired_order]  #reorder rdm

## freesort-first group
df_causal_rdm_freesort_first <- df_freesort_first %>%
  mutate(item_pairs = str_c(itemA, itemB, sep = "-")) %>%
  group_by(item_pairs) %>%
  mutate(mean_causal = mean(causal_distance, na.rm = TRUE)) %>%
  ungroup() %>%
  distinct(item_pairs, .keep_all = TRUE) %>%
  select(itemA, itemB, mean_causal) %>%
  pivot_wider(names_from = itemB, values_from = mean_causal) %>%
  column_to_rownames("itemA")

df_causal_rdm_freesort_first <- df_causal_rdm_freesort_first[desired_order, desired_order]  #reorder rdm

# Getting freesort RDMs

## causation-first group
df_freesort_rdm_causation_first <- df_causal_first %>%
  mutate(item_pairs = str_c(itemA, itemB, sep = "-")) %>%
  group_by(item_pairs) %>%
  mutate(mean_euclidean = mean(freesort_distance, na.rm = TRUE)) %>%
  ungroup() %>%
  distinct(item_pairs, .keep_all = TRUE) %>%
  select(itemA, itemB, mean_euclidean) %>%
  pivot_wider(names_from = itemB, values_from = mean_euclidean) %>%
  column_to_rownames("itemA")

df_freesort_rdm_causation_first <- df_freesort_rdm_causation_first[desired_order, desired_order]  #reorder rdm

## freesort-first group
df_freesort_rdm_freesort_first <- df_freesort_first %>%
  mutate(item_pairs = str_c(itemA, itemB, sep = "-")) %>%
  group_by(item_pairs) %>%
  mutate(mean_euclidean = mean(freesort_distance, na.rm = TRUE)) %>%
  ungroup() %>%
  distinct(item_pairs, .keep_all = TRUE) %>%
  select(itemA, itemB, mean_euclidean) %>%
  pivot_wider(names_from = itemB, values_from = mean_euclidean) %>%
  column_to_rownames("itemA")

df_freesort_rdm_freesort_first <- df_freesort_rdm_freesort_first[desired_order, desired_order]  #reorder rdm


#flatten the upper tris, for correlation
flatten_rdm <- function(matrix) {
  matrix[upper.tri(matrix)]
}

# Extract upper triangles
causation_rdm_correlation <- cor(flatten_rdm(as.matrix(df_causal_rdm_causation_first)),
                                 flatten_rdm(as.matrix(df_causal_rdm_freesort_first)),
                                 method = "pearson")

freesort_rdm_correlation <- cor(flatten_rdm(as.matrix(df_freesort_rdm_causation_first)),
                                flatten_rdm(as.matrix(df_freesort_rdm_freesort_first)),
                                method = "pearson")

print(causation_rdm_correlation)
print(freesort_rdm_correlation)

#compute correlation difference
correlation_difference <- causation_rdm_correlation - freesort_rdm_correlation

```

## causal-freesort vs causal freesort correlation

```{r}
#get upper tris
causation_first_rdm_correlation <- cor(flatten_rdm(as.matrix(df_causal_rdm_causation_first)),
                                 flatten_rdm(as.matrix(df_freesort_rdm_causation_first)),
                                 method = "pearson")

freesort_first_rdm_correlation <- cor(flatten_rdm(as.matrix(df_causal_rdm_freesort_first)),
                                flatten_rdm(as.matrix(df_freesort_rdm_freesort_first)),
                                method = "pearson")

print(causation_first_rdm_correlation)
print(freesort_first_rdm_correlation)

#compute correlation difference
correlation_difference_orders <- freesort_first_rdm_correlation - causation_first_rdm_correlation 
correlation_difference_orders

fisher_z <- function(r) {
  return(0.5 * log((1 + r) / (1 - r)))
}

#convert correlations to Fisher's z
z1 <- fisher_z(causation_first_rdm_correlation)
z2 <- fisher_z(freesort_first_rdm_correlation)

N1 <- nrow(df_causal_rdm_causation_first)  
N2 <- nrow(df_causal_rdm_freesort_first)  

SE <- sqrt(1 / (N1 - 3) + 1 / (N2 - 3))

z_diff <- (z2 - z1) / SE #z score for the difference
p_value <- 2 * (1 - pnorm(abs(z_diff)))

print(paste("Z-score:", z_diff))
print(paste("P-value:", p_value))
```



```{r}

```

