---
title: "study1 preprocessing"
output: 
  html_document:
      toc: true
      toc_float: true
      toc_depth: 3
      number_sections: true
date: "2024-12-10"
author: anonymous
---

This script does the following:  
1. Takes raw data `study1_data_raw.csv`. The codebook for the raw data is available at `data/study1_data_raw_codebook`.  
2. Parses it into a combined (i.e. freesort + causation) data frame and saves into csv `data_processed_step1.csv`. The codebook for this data frame is available at `/snapshots_study1_preprocessing_analysis/data_processed_step1_codebook`.  
3. Generates model predictions (aka theoretical RDM data) and saves into csv `data_model_predictions.csv`.  
4. Plots RDMs and saves them into `figures/theoretical-rdms`. 

Study description:    
- Study 1 was ran on 12/13/24 on Connect CloudResearch, N = 50 participants.  
- Subject selection criteria: 95% approval rate, minimum 10 previous submissions, and fluent in English. 

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
pacman::p_load(tidyverse, here, corrplot, jsonlite)
sessionInfo()

library(jsonlite)
```

# Read data

```{r}
base_path <- here("code", "study1")
data_path <- paste0(base_path, "/data/study1_data_raw.csv")
d <- read_csv(data_path)
```

# Set variables

```{r}
items <- c("choose what to do", "reach for something", "feel scared",  #items in exact order of img1.png to img15.png used in sort task
           "experience pain", "get sick", "get tired", "become hungry", 
           "see something", "think about something", "hear something", 
           "remember something", "jump up and down", "kick something", 
           "take a walk", "sit down")

desired_order <- c("see something", "hear something", "choose what to do", #ordering by domain
                   "remember something", "think about something", "reach for something", 
                   "sit down", "jump up and down", "kick something", "take a walk", 
                   "get tired", "become hungry", "feel scared", "experience pain", "get sick")

#paths: data outputs
rdm_figures_outputs_path <- paste0(base_path, "/figures/theoretical-rdms/")
combined_output_csv <- "data_processed_step1"
snapshots_path <- paste0(base_path, "/snapshots_study1_preprocessing_analysis")
combined_output_path <- paste0(snapshots_path, "/", combined_output_csv, ".csv") 
model_predictions_output_path <- paste0(snapshots_path, "/data_model_predictions.csv") 

```


# Parse data frames

## Parse Sorting Task data

```{r}
data_freesort <- d %>%
  filter(trial_type == "free-sort") %>% 
  select(subject_id, trial_index, final_locations) %>% 
  mutate(parsed_location = map(final_locations, ~ fromJSON(.))) %>% #parse JSON
  unnest(parsed_location) %>% #unnest the list column
  mutate(
    img_number = as.numeric(str_extract(src, "\\d+")), #etract image numbers from `src`
    item = items[img_number] #map image numbers to item labels
  ) %>%
  select(-c("final_locations", "src", "img_number")) %>% 
  mutate(trial_type = "free_sort") %>% 
  relocate(trial_type, .after = subject_id) %>% 
  rename(original_trial_index = trial_index)

#write.csv(data_freesort, file = freesort_output_path, row.names = F)
```

## Parse Causal Task data

```{r}
response_data <- d %>%
  filter(trial_type == "survey-slider") %>% 
  select(subject_id, trial_index, response) %>%
  mutate(response = map(response, ~ tryCatch(
    fromJSON(.) %>%
      enframe(name = "question_id", value = "response"),
    error = function(e) tibble(question_id = character(), response = character())
  ))) %>%
  unnest(response) %>% 
  mutate(
    response = unlist(response), 
    response = as.numeric(response), 
    question_index = str_remove(question_id, "Q") 
  ) %>%
  select(subject_id, trial_index, question_index, response)

metadata_data <- d %>%
  select(subject_id, trial_index, `0`:`14`, `15`) %>% 
  pivot_longer(
    cols = -c(subject_id, trial_index),  
    names_to = "question_index",      
    values_to = "metadata"            
  ) %>% 
  mutate(
    parsed_metadata = map(metadata, ~ {
      if (is.na(.x)) {
        tibble() # Handle NA metadata
      } else {
        fromJSON(.x, simplifyVector = FALSE) %>%
          map(~ if (is.null(.)) NA else .) %>%  #replace NULL with NA
          as_tibble()
      }
    })
  ) %>% 
  unnest(parsed_metadata) %>%    
  select(-metadata)           

#merge the response_data and metadata_data into data_causation
data_causation <- response_data %>%
  inner_join(metadata_data, by = c("subject_id", "trial_index", "question_index")) %>% 
  mutate(trial_type = "causation") %>% 
  relocate(trial_type, .after = subject_id) %>%
  rename(original_trial_index = trial_index, 
         original_question_index = question_index) %>% 
  group_by(subject_id, trial_type) %>%
  mutate(trial_index = dense_rank(original_trial_index),
         question_index = row_number()) %>%
  ungroup() %>% 
  relocate(c("trial_index", "question_index"), .after = "trial_type") %>% 
  relocate(c("original_trial_index", "original_question_index"), .after = "correct_response")

```

## Exclusions

- In Causal Task we have 4 attention checks at pages 4, 8, 10 and 13. No attention checks in Sorting Task 
- We are excluding subjects who fail at least 2 attention checks (Other possible criteria: time taken to complete study)

```{r}
data_causation <- data_causation %>%
  mutate(attention_check_id = case_when(
    attention_check == TRUE & trial_index == 4  ~ 1, 
    attention_check == TRUE & trial_index == 8  ~ 2, 
    attention_check == TRUE & trial_index == 10 ~ 3, 
    attention_check == TRUE & trial_index == 13 ~ 4, 
    TRUE ~ NA_integer_ #non-attention check rows
  )) %>%
  mutate(correct_response = case_when(
    attention_check == TRUE & trial_index == 4  ~ 0, #"Move slider all the way to left"
    attention_check == TRUE & trial_index == 8  ~ 75,  #"Move slider to 'Maybe yes'"
    attention_check == TRUE & trial_index == 10 ~ 100, #"Move slider all the way to right"
    attention_check == TRUE & trial_index == 13 ~ 0,   #"Move slider to 'Definitely not'"
    TRUE ~ correct_response # Retain existing values for non-attention check rows
  )) %>%
  mutate(correct = case_when(
    attention_check == TRUE & trial_index == 4  & response >= 0 & response <= 5   ~ TRUE, 
    attention_check == TRUE & trial_index == 8  & response >= 70 & response <= 80 ~ TRUE, 
    attention_check == TRUE & trial_index == 10 & response >= 0 & response <= 5 ~ TRUE, #3 and 4 are switched for some reason
    attention_check == TRUE & trial_index == 13 & response >= 95 & response <= 100   ~ TRUE,
    attention_check == TRUE ~ FALSE, #outside acceptable range
    TRUE ~ NA #non-attn check rows
  ))

excluded_participants <- data_causation %>%
  filter(attention_check == TRUE) %>% 
  group_by(subject_id) %>% 
  summarize(
    num_incorrect = sum(correct == FALSE, na.rm = TRUE), 
    .groups = "drop"
  ) %>%
  filter(num_incorrect >= 2) %>% #identify participants with x or more incorrect (4 subjects had 1 incorrect attn check)
  pull(subject_id) 

data_causation <- data_causation %>%
  filter(!subject_id %in% excluded_participants)

```

## Add trial order column

```{r}
#Combine trial types to compare their starting original_trial_index
trial_order_info <- bind_rows(
  data_causation %>% distinct(subject_id, trial_type, original_trial_index),
  data_freesort %>% distinct(subject_id, trial_type, original_trial_index)
) %>%
  group_by(subject_id) %>%
  summarize(
    trial_type_order = case_when(
      min(original_trial_index[trial_type == "causation"]) < min(original_trial_index[trial_type == "free_sort"]) ~ 
        list(c(causation = "first", free_sort = "second")),
      TRUE ~ list(c(causation = "second", free_sort = "first"))
    )
  ) %>%
  unnest_wider(trial_type_order)

#Add trial_type_order to data_causation
data_causation <- data_causation %>%
  left_join(
    trial_order_info %>% select(subject_id, causation),
    by = "subject_id"
  ) %>%
  rename(trial_type_order = causation) %>%
  relocate(trial_type_order, .after = trial_type)

# Add trial_type_order to data_freesort
data_freesort <- data_freesort %>%
  left_join(
    trial_order_info %>% select(subject_id, free_sort),
    by = "subject_id"
  ) %>%
  rename(trial_type_order = free_sort) %>%
  relocate(trial_type_order, .after = trial_type)

```

## Create combined data frame

```{r}
data_combined <- bind_rows(
  data_causation,
  data_freesort
)

assign(combined_output_csv, data_combined)

write.csv(data_combined, combined_output_path, row.names = F)

```

# Generate theoretical RDM data 

These are the model predictions for the 4 hypotheses

```{r}
d_theories_first3 <- expand.grid(itemA = desired_order, itemB = desired_order) %>% 
  filter(itemA != itemB) %>%  # Remove self-pairs
  
  #theory 1: mind, body, action
  mutate(MindBodyAction_itemA = case_when(itemA == "hear something" ~ "mind", #labels for first item of pair
                               itemA == "choose what to do" ~ "mind",
                               itemA == "remember something" ~ "mind",
                               itemA == "think about something" ~ "mind",
                               itemA == "see something" ~ "mind",
                               itemA == "reach for something" ~ "act",
                               itemA == "sit down" ~ "act", 
                               itemA == "jump up and down" ~ "act",
                               itemA == "kick something" ~ "act",
                               itemA == "take a walk" ~ "act",
                               itemA == "get tired" ~ "bio", 
                               itemA == "become hungry" ~ "bio",
                               itemA == "feel scared"~ "bio",
                               itemA == "experience pain"~ "bio",
                               itemA == "get sick"~ "bio"),
         MindBodyAction_itemB = case_when(itemB == "hear something" ~ "mind", #labels for second item of pair
                               itemB == "choose what to do" ~ "mind",
                               itemB == "remember something" ~ "mind",
                               itemB == "think about something" ~ "mind",
                               itemB == "see something" ~ "mind",
                               itemB == "reach for something" ~ "act",
                               itemB == "sit down" ~ "act", 
                               itemB == "jump up and down" ~ "act",
                               itemB == "kick something" ~ "act",
                               itemB == "take a walk" ~ "act",
                               itemB == "get tired" ~ "bio", 
                               itemB == "become hungry" ~ "bio",
                               itemB == "feel scared"~ "bio",
                               itemB == "experience pain"~ "bio",
                               itemB == "get sick"~ "bio")) %>% 
  
  #theory 2: fine-grained, mba: perceptual, cognition, object & non-object directed, phasic & non phasic changes
   mutate(PerCogOdActSpActBstimBnstim_itemA = case_when(itemA == "hear something" ~ "per",
                               itemA == "see something" ~ "per",
                               itemA == "choose what to do" ~ "cog",
                               itemA == "remember something" ~ "cog",
                               itemA == "think about something" ~ "cog",
                               itemA == "reach for something" ~ "odAct",
                               itemA == "sit down" ~ "SpAct", 
                               itemA == "jump up and down" ~ "SpAct",
                               itemA == "kick something" ~ "odAct",
                               itemA == "take a walk" ~ "SpAct",
                               itemA == "get tired" ~ "BnS", 
                               itemA == "become hungry" ~ "BnS",
                               itemA == "feel scared"~ "BS",
                               itemA == "experience pain"~ "BS",
                               itemA == "get sick"~ "BnS",),
         PerCogOdActSpActBstimBnstim_itemB = case_when(itemB == "hear something" ~ "per",
                               itemB == "see something" ~ "per",
                               itemB == "choose what to do" ~ "cog",
                               itemB == "remember something" ~ "cog",
                               itemB == "think about something" ~ "cog",
                               itemB == "reach for something" ~ "odAct",
                               itemB == "sit down" ~ "SpAct", 
                               itemB == "jump up and down" ~ "SpAct",
                               itemB == "kick something" ~ "odAct",
                               itemB == "take a walk" ~ "SpAct",
                               itemB == "get tired" ~ "BnS", 
                               itemB == "become hungry" ~ "BnS",
                               itemB == "feel scared"~ "BS",
                               itemB == "experience pain"~ "BS",
                               itemB == "get sick"~ "BnS")) %>% 

      #theory 3: (2 groups) physical and non-physical/psychological events:
      mutate(PhysicalEthereal_itemA = case_when(itemA == "hear something" ~ "ethereal",
                               itemA == "see something" ~ "ethereal",
                               itemA == "choose what to do" ~ "ethereal",
                               itemA == "remember something" ~ "ethereal",
                               itemA == "think about something" ~ "ethereal",
                               itemA == "reach for something" ~ "phys",
                               itemA == "sit down" ~ "phys", 
                               itemA == "jump up and down" ~ "phys",
                               itemA == "kick something" ~ "phys",
                               itemA == "take a walk" ~ "phys",
                               itemA == "get tired" ~ "phys", 
                               itemA == "become hungry" ~ "phys",
                               itemA == "feel scared"~ "phys",
                               itemA == "experience pain"~ "phys",
                               itemA == "get sick"~ "phys"),
             PhysicalEthereal_itemB = case_when(itemB == "hear something" ~ "ethereal",
                               itemB == "see something" ~ "ethereal",
                               itemB == "choose what to do" ~ "ethereal",
                               itemB == "remember something" ~ "ethereal",
                               itemB == "think about something" ~ "ethereal",
                               itemB == "reach for something" ~ "phys",
                               itemB == "sit down" ~ "phys", 
                               itemB == "jump up and down" ~ "phys",
                               itemB == "kick something" ~ "phys",
                               itemB == "take a walk" ~ "phys",
                               itemB == "get tired" ~ "phys", 
                              itemB == "become hungry" ~ "phys",
                               itemB == "feel scared"~ "phys",
                               itemB == "experience pain"~ "phys",
                               itemB == "get sick"~ "phys")) %>% 
  
  #scoring the items
  mutate("theor1_Mind-Body-Action" = ifelse(MindBodyAction_itemA == MindBodyAction_itemB, 0, 1),
         "theor2_PerCog-OdActSpAct-BodyStimBodynoStim" = ifelse(PerCogOdActSpActBstimBnstim_itemA == PerCogOdActSpActBstimBnstim_itemB, 0, 1),
         "theor3_Physical-Ethereal" = ifelse(PhysicalEthereal_itemA == PhysicalEthereal_itemB, 0, 1)) 

# write.csv(d_theories_first3, theories_output_path, row.names = F)

```

## Add USE embeddings

```{r}
#load USE embeddings
embedding_df <- readRDS(here(base_path, "embeddings", "item_embeddings.rds")) 

d_theories <- d_theories_first3 %>% 
  select(-c("MindBodyAction_itemA", "MindBodyAction_itemB", 
            "PerCogOdActSpActBstimBnstim_itemA","PerCogOdActSpActBstimBnstim_itemB", 
            "PhysicalEthereal_itemA", "PhysicalEthereal_itemB"))

#merge embeddings to original data frame
d_theories_with_embeddings <- d_theories %>%
  left_join(embedding_df, by = c("itemA" = "item")) %>%
  rename(embeddingA = embedding) %>%
  left_join(embedding_df, by = c("itemB" = "item")) %>%
  rename(embeddingB = embedding)

#compute cosine similarity for each row
d_all_theories <- d_theories_with_embeddings %>%
  rowwise() %>% 
  mutate(cosine_similarity = sum(unlist(embeddingA) * unlist(embeddingB)) /
                 (sqrt(sum(unlist(embeddingA)^2)) * sqrt(sum(unlist(embeddingB)^2)))) %>%
  ungroup() %>% 
  rowwise() %>% 
  mutate(cosine_similarity = (1 - cosine_similarity) / 2) %>% #turn c.s. to a disimilarity measure
  ungroup() %>% 
  select(-c("embeddingA", "embeddingB")) %>% 
  rename("theor4_cosine_similarity" = cosine_similarity) %>% 
  ungroup() 

write.csv(d_all_theories, model_predictions_output_path, row.names = F)

```

# Plot theoretical RDMS

```{r}
plot_rdm_theory <- function(data, column_name, desired_order, plot_title) {

  df_matrix <- data %>%
    mutate(
      itemA = factor(itemA, levels = desired_order),
      itemB = factor(itemB, levels = desired_order)
    ) %>%
    select(itemA, itemB, all_of(column_name)) %>%
    pivot_wider(names_from = itemB, values_from = all_of(column_name)) %>%
    column_to_rownames("itemA")
  
  df_matrix <- df_matrix[desired_order, desired_order]
  diag(df_matrix) <- NA 
  
  #min-max normalize
  min_val <- min(df_matrix, na.rm = TRUE)
  max_val <- max(df_matrix, na.rm = TRUE)
  df_matrix <- (df_matrix - min_val) / (max_val - min_val)

  custom_palette <- colorRampPalette(c("red", "white", "blue"))

  corrplot(
    as.matrix(df_matrix), method = "color", is.corr = FALSE,
    col = custom_palette(200),
    tl.col = "black", 
    #title = plot_title,
    title = "",
    cex.main = 3, #title text size
    tl.pos = "n", #remove labels
    mar = c(0, 0, 2, 0), addgrid.col = "darkgray",
    na.label = "square",       
    na.label.col = "darkgrey",
    cl.offset = 1.0,  
    cl.cex = 2, #legend label text size
    cl.ratio = 0.22, #widen the legend bar: change this to push words far away
  )
}

mba_order <- c("see something", "hear something", "choose what to do", 
               "remember something", "think about something", "reach for something", 
               "sit down", "jump up and down", "kick something", "take a walk", 
               "get tired", "become hungry", "feel scared", "experience pain", "get sick")

perccog_order <- c("see something", "hear something", "choose what to do", 
                   "remember something", "think about something", "reach for something", 
                   "kick something", "sit down", "jump up and down",  "take a walk", 
                   "get tired", "become hungry", "get sick", "feel scared", "experience pain")

physethereal_order <- c("see something", "hear something", "choose what to do", 
                        "remember something", "think about something", "reach for something", 
                        "kick something", "sit down", "jump up and down",  "take a walk", 
                        "get tired", "become hungry", "get sick", "feel scared", "experience pain")

plot_rdm_theory(d_all_theories, "theor4_cosine_similarity", mba_order, "Cosine Distance")
plot_rdm_theory(d_all_theories, "theor1_Mind-Body-Action", mba_order, "3 categories: MBA")
plot_rdm_theory(d_all_theories, "theor2_PerCog-OdActSpAct-BodyStimBodynoStim", perccog_order, "6 categories: PCOS")
plot_rdm_theory(d_all_theories, "theor3_Physical-Ethereal", physethereal_order, "2 categories: PE")

# save them
#output_folder <- paste0(here(rdm_outputs_path), "/figures/theoretical-rdms/") -> delete this

theories <- list(
  list(column_name = "theor4_cosine_similarity", order = mba_order, title = "Cosine Distance"),
  list(column_name = "theor1_Mind-Body-Action", order = mba_order, title = "3 categories: MBA"),
  list(column_name = "theor2_PerCog-OdActSpAct-BodyStimBodynoStim", order = perccog_order, title = "6 categories: PCOS"),
  list(column_name = "theor3_Physical-Ethereal", order = physethereal_order, title = "2 categories: PE")
)

for (i in seq_along(theories)) {
  theory <- theories[[i]]
  output_file <- paste0(rdm_figures_outputs_path, "theory_", i, ".pdf")
  pdf(file = output_file, width = 8, height = 6) #open PDF device
  plot_rdm_theory(
    data = d_all_theories,
    column_name = theory$column_name,
    desired_order = theory$order,
    plot_title = theory$title
  )
  dev.off() #close PDF device
}

```

# Demographics

```{r}
df_demographics <- d %>% 
  filter(trial_id == "demographics_survey") %>% 
  select("subject_id", "trial_id","feedback", "age", "ethnicity", "gender", "race") 

summary(df_demographics$age)
table(df_demographics$gender)
table(df_demographics$ethnicity)
table(df_demographics$race)

```



```{r}

```


