---
title: "study 2"
output: 
  html_document:
      toc: true
      toc_float: true
      toc_depth: 3
      number_sections: true
date: "2025-1-10"
author: anonymous
---

This is the analysis script for study 2 of the project "Adults hold a causal explanatory framework that connects understanding of other people’s minds, actions and bodies" by (authors redacted).

Study 2 had N = 100, with random assignment to two conditions: Intervention and Trait Inference, with 15 trials per condition.

The question: are subjects in the Intervention condition more likely to select the causal item, and subjects in the Inference condition more likely to select the similar item.

This script takes as input `study2_data_raw.csv` and `data_freesort_minus_causals.csv`, and plots Supplementary Table 1 in the SI materials, Figure 2 from the main manuscript, and runs statistical models to predict variability in responses from causal and freesort distances.

# Setup

## Set options

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
pacman::p_load(tidyverse, 
               lme4, 
               here, 
               patchwork, 
               cowplot, 
               viridis, 
               effects, 
               gt,
               webshot2,
               knitr,
               performance)
sessionInfo()

```

## Set variables

```{r, warnings = FALSE}
# variables
desired_order <- c("see something", "hear something", "choose what to do", 
                   "remember something", "think about something", "reach for something", 
                   "sit down", "jump up and down", "kick something", "take a walk", 
                   "get tired", "become hungry", "feel scared", "experience pain", "get sick")

# paths
## script inputs
base_path <- here("code")
data_path <- paste0(base_path, "/study2/data/study2_data_raw.csv")
snapshots_input_path <- paste0(base_path, "/study1/snapshots_study1_final_analysis")
freesort_and_causal_distances_df_path <- paste0(snapshots_input_path, "/freesort_minus_causals.csv")

```

## Read data

```{r}
d <- read.csv(data_path)

```


## Demographics

```{r}
df_demographics <- d %>% 
  filter(trial_id == "demographics_survey") %>% 
  select("subject_id", "trial_id", 
         "feedback", "age", "ethnicity", "gender", "race") 

summary(df_demographics$age)
table(df_demographics$gender)
table(df_demographics$ethnicity)
table(df_demographics$race)

```


## Wrangle data frame

```{r}
 subjects_to_exclude <- d %>%
  filter(condition == "attention_check") %>% 
  rowwise() %>% 
  mutate(correct_attn_check = case_when(
    grepl(correct_response_attn, selected_choice) ~ "yes",
    TRUE ~ "no"
  )) %>% 
  filter(correct_attn_check == "no") %>% 
  pull(subject_id) #1 subject: subject 11

dat <- d %>% 
  filter(!subject_id %in% subjects_to_exclude) %>% #exclude 1 subject who failed attn check
  filter(condition %in% c("intervention", "similarity")) %>%
  mutate(response = as.factor(response)) %>% 
  mutate(condition = as.factor(condition)) %>% 
  rowwise() %>%
  mutate(response = case_when(
    grepl(causal_itemA, selected_choice) ~ "causal",
    grepl(similarity_itemA, selected_choice) ~ "similar",
    TRUE ~ "other"
  )) %>% 
  mutate(condition = case_when(
    condition == "similarity" ~ "inference",
    TRUE ~ condition
  )) %>% 
  ungroup() %>% 
  select(subject_id, condition, itemB, causal_itemA, similarity_itemA, response) %>% 
  mutate(causal_itemA = factor(causal_itemA, levels = desired_order),
         similarity_itemA = factor(similarity_itemA, levels = desired_order),
         itemB = factor(itemB, levels = desired_order)) %>% 
  mutate(domain_itemB = case_when(itemB == "hear something" ~ "mind", #labels for first item of pair
                                 itemB == "choose what to do" ~ "mind",
                                 itemB == "remember something" ~ "mind",
                                 itemB == "think about something" ~ "mind",
                                 itemB == "see something" ~ "mind",
                                 itemB == "reach for something" ~ "action",
                                 itemB == "sit down" ~ "action", 
                                 itemB == "jump up and down" ~ "action",
                                 itemB == "kick something" ~ "action",
                                 itemB == "take a walk" ~ "action",
                                 itemB == "get tired" ~ "bio", 
                                 itemB == "become hungry" ~ "bio",
                                 itemB == "feel scared"~ "bio",
                                 itemB == "experience pain"~ "bio",
                                 itemB == "get sick"~ "bio"),
         domain_causalitemA = case_when(causal_itemA == "hear something" ~ "mind", 
                                 causal_itemA == "choose what to do" ~ "mind",
                                 causal_itemA == "remember something" ~ "mind",
                                 causal_itemA == "think about something" ~ "mind",
                                 causal_itemA == "see something" ~ "mind",
                                 causal_itemA == "reach for something" ~ "action",
                                 causal_itemA == "sit down" ~ "action", 
                                 causal_itemA == "jump up and down" ~ "action",
                                 causal_itemA == "kick something" ~ "action",
                                 causal_itemA == "take a walk" ~ "action",
                                 causal_itemA == "get tired" ~ "bio", 
                                 causal_itemA == "become hungry" ~ "bio",
                                 causal_itemA == "feel scared"~ "bio",
                                 causal_itemA == "experience pain"~ "bio",
                                 causal_itemA == "get sick"~ "bio"),
          domain_similarityitemA = case_when(similarity_itemA == "hear something" ~ "mind", 
                                 similarity_itemA == "choose what to do" ~ "mind",
                                 similarity_itemA == "remember something" ~ "mind",
                                 similarity_itemA == "think about something" ~ "mind",
                                 similarity_itemA == "see something" ~ "mind",
                                 similarity_itemA == "reach for something" ~ "action",
                                 similarity_itemA == "sit down" ~ "action", 
                                 similarity_itemA == "jump up and down" ~ "action",
                                 similarity_itemA == "kick something" ~ "action",
                                 similarity_itemA == "take a walk" ~ "action",
                                 similarity_itemA == "get tired" ~ "bio", 
                                 similarity_itemA == "become hungry" ~ "bio",
                                 similarity_itemA == "feel scared"~ "bio",
                                 similarity_itemA == "experience pain"~ "bio",
                                 similarity_itemA == "get sick"~ "bio")) 


```

# Plot table of distances

## make df of choices

```{r}
design_pairs <- dat %>% #start with raw data containing items
  select(itemB, causal_itemA, similarity_itemA) %>%
  unique() %>%
  left_join(
    read.csv(freesort_and_causal_distances_df_path) %>% #grab causal item distances
      mutate(
        itemB = factor(itemB, levels = desired_order),
        itemA = factor(itemA, levels = desired_order)
      ) %>%
      select(itemA, itemB, mean_causal_distance, mean_freesort_distance),
    by = c("itemB", "causal_itemA" = "itemA")
  ) %>%
  left_join(
    read.csv(freesort_and_causal_distances_df_path) %>% #grab similar item distances
      mutate(
        itemB = factor(itemB, levels = desired_order),
        itemA = factor(itemA, levels = desired_order)
      ) %>%
      select(itemA, itemB, mean_causal_distance, mean_freesort_distance),
    by = c("itemB", "similarity_itemA" = "itemA"),
    suffix = c("_causal", "_similar")
  ) %>%
  distinct(itemB, .keep_all = TRUE) %>%
  mutate(
    mean_causal_distance_causal = round(mean_causal_distance_causal, 2),
    mean_freesort_distance_causal = round(mean_freesort_distance_causal, 2),
    mean_causal_distance_similar = round(mean_causal_distance_similar, 2),
    mean_freesort_distance_similar = round(mean_freesort_distance_similar, 2)
  ) %>%
  select(
    target_item = itemB,
    causal_choice = causal_itemA,
    causal_distance_causal_choice = mean_causal_distance_causal,
    freesort_distance_causal_choice = mean_freesort_distance_causal,
    similar_choice = similarity_itemA,
    causal_distance_similar_choice = mean_causal_distance_similar,
    freesort_distance_similar_choice = mean_freesort_distance_similar
  ) %>%
  arrange(target_item)

```

## figure for paper: table with domain words highlighted

```{r}
pairs_table <- design_pairs %>%
  mutate(
    causal_choice = paste0(causal_choice, " (", causal_distance_causal_choice, ", ", freesort_distance_causal_choice, ")"),
    similar_choice = paste0(similar_choice, " (", causal_distance_similar_choice, ", ", freesort_distance_similar_choice, ")"),
    group = case_when(
      row_number() %in% 1:5 ~ "mind",
      row_number() %in% 6:10 ~ "action",
      row_number() %in% 11:15 ~ "body"
    ),
    #inline styles to highlight text for 'target_item' based on group
    target_item = case_when(
      group == "mind" ~ paste0("<span style='color: #f7bf2a; font-weight: bold;'>", target_item, "</span>"),
      group == "action" ~ paste0("<span style='color: #69ad44; font-weight: bold;'>", target_item, "</span>"),
      group == "body" ~ paste0("<span style='color: #f36b2d; font-weight: bold;'>", target_item, "</span>"),
      TRUE ~ target_item
    )
  ) %>%
  #replace group labels with blanks after the first row of each group
  group_by(group) %>%
  mutate(group = ifelse(row_number() == 1, group, "")) %>%
  ungroup() %>%
  select(group, target_item, causal_choice, similar_choice)  # Keep the domain column


df_table <- pairs_table %>%
  gt() %>%
  tab_header(
    title = "Target and Choice Items"
  ) %>%
  cols_label(
    group = "Domain",
    target_item = "Target Item",
    causal_choice = md("Causal Choice<br>(Causal Distance, Freesort Distance)"),
    similar_choice = md("Similar Choice<br>(Causal Distance, Freesort Distance)")
  ) %>%
  fmt_markdown(columns = target_item) %>%  # Render markdown for inline styles
  fmt_markdown(columns = c(causal_choice, similar_choice)) %>%
  #custom styles for domain (group) column
  tab_style(
    style = list(
      cell_text(color = "#f7bf2a", weight = "bold")  
    ),
    locations = cells_body(
      columns = group,
      rows = group == "mind"
    )
  ) %>%
  tab_style(
    style = list(
      cell_text(color = "#69ad44", weight = "bold")
    ),
    locations = cells_body(
      columns = group,
      rows = group == "action"
    )
  ) %>%
  tab_style(
    style = list(
      cell_text(color = "#f36b2d", weight = "bold") 
    ),
    locations = cells_body(
      columns = group,
      rows = group == "body"
    )
  ) %>%
  tab_options(
    table.font.names = "Arial"
  )

gtsave(df_table, "figures/pairs-table-words.png")

```


# Barplot: condition effect

## summarize data

```{r}
dat_summarized <- dat %>%
  mutate(response = fct_relevel(response, c("similar", "causal"))) %>% 
  group_by(condition) %>%
  mutate(total = n()) %>% 
  group_by(condition, response) %>%
  summarize(
    count = n(),  
    total = first(total), 
    mean_proportion = count / total, 
    se = sqrt(mean_proportion * (1 - mean_proportion) / total),  
    .groups = "drop")

```

## plot stacked barplot

```{r}

#stacked
(fig_result_summary <- ggplot(dat_summarized, aes(x = condition, y = mean_proportion, fill = response)) +
  geom_bar(stat = "identity", colour = "black") +
  scale_fill_manual(values = c("grey80", "#000000")) + 
  geom_errorbar(data = filter(dat_summarized, response == "causal"), aes(ymax = mean_proportion + se,
                                                                         ymin = mean_proportion - se),
                stat = "identity", position = "identity", width = 0.3) +
  theme_cowplot() +
  labs(y = "Proportion"))

```

# Barplot: item level, stacked

## summarize df

```{r, fig.width = 12}
dat_summarized_by_item <- dat %>% 
  mutate(response = fct_relevel(response, c("similar", "causal"))) %>% 
  group_by(condition, itemB) %>%
  mutate(total = n()) %>%  #get total responses per condition
  group_by(condition, itemB, response) %>%
  mutate(
    count = n(),  #count responses for each response type
    total = first(total),  #total responses for the condition
    mean_proportion = count / total,  #prop response within condition
    se = sqrt(mean_proportion * (1 - mean_proportion) / total)) %>% 
  distinct(condition, itemB, .keep_all = TRUE) %>% 
  ungroup()

```

## figure for paper

```{r}
itemB_colors <- dat_summarized_by_item %>%
  distinct(itemB, domain_itemB) %>%
  mutate(color = case_when(
    domain_itemB == "bio" ~ "darkgreen",
    domain_itemB == "mind" ~ "blue",
    domain_itemB == "action" ~ "red"
  )) %>%
  pull(color, name = itemB)

p <- dat_summarized_by_item %>% 
  ggplot(aes(x =itemB, y = mean_proportion, fill = response)) +
  geom_bar(stat = "identity") +
  geom_bar(stat = "identity", colour = "black") +
  geom_errorbar(data = filter(dat_summarized_by_item, response == "causal"), aes(ymax = mean_proportion + se,
                                                                         ymin = mean_proportion - se),
                stat = "identity", position = "identity", width = 0.3) +
  scale_fill_manual(values = c("grey80", "#000000")) + 
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = itemB_colors[levels(dat_summarized_by_item$itemB)])) +
  facet_wrap(~condition) 

p_alt <- dat_summarized_by_item %>% 
  ggplot(aes(x =condition, y = mean_proportion, fill = response)) +
  geom_bar(stat = "identity") +
  geom_bar(stat = "identity", colour = "black") +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  geom_errorbar(data = filter(dat_summarized_by_item, response == "causal"), aes(ymax = mean_proportion + se,
                                                                         ymin = mean_proportion - se),
                stat = "identity", position = "identity", width = 0.3) +
  scale_fill_manual(values = c("grey80", "#000000")) +
  theme_cowplot() +
  # theme(axis.text.x = element_text(angle = 45, hjust = 1, color = itemB_colors[levels(dat_summarized_by_item$itemB)])) +
  theme(axis.text.x = element_text(size = 13),
        axis.text.y = element_text(size = 13),
        legend.position = "bottom") +
  labs(y = "proportion") +
  facet_wrap(~itemB, nrow = 3) 

p_alt

ggsave(
  filename = "figures/p_study2_item-level_stacked.pdf",
  plot = p_alt,          
  device = "pdf",           
  width = 12,               
  height = 8,             
  units = "in"              
)

```

# Modelling

## prepare data frames

```{r}
d_model <- dat %>% 
  mutate(response = as.factor(response)) %>% 
  mutate(subject_id = as.factor(subject_id)) %>% 
  mutate(responses = ifelse(response == "causal", 1, 0)) %>% 
  mutate(condition = as.factor(condition)) %>%
  unite(col = "trial_ID", c(itemB, similarity_itemA, causal_itemA), sep = "_", remove = FALSE)
  
d_freesort_and_causal_distances <- read.csv(freesort_and_causal_distances_df_path) %>% 
  mutate(itemB = factor(itemB, levels = desired_order)) %>% 
  mutate(itemA = factor(itemA, levels = desired_order)) 

intervention_condition <- d_model %>% 
  # select(-similarity_itemA) %>% # why - let's study both?
    filter(condition == "intervention") %>%
    left_join(d_freesort_and_causal_distances %>% 
                select(itemA, itemB, mean_causal_distance, mean_freesort_distance, freesort_minus_causal) %>% 
                rename("causal_itemA" = itemA,
                       "causal_itemA_causal_distance" = mean_causal_distance,
                       "causal_itemA_freesort_distance" = mean_freesort_distance,
                       "causal_itemA_freesort_minus_causal" = freesort_minus_causal),
              by = c("itemB", "causal_itemA")) %>%
    left_join(d_freesort_and_causal_distances %>% 
                select(itemA, itemB, mean_causal_distance, mean_freesort_distance, freesort_minus_causal) %>% 
                rename("similarity_itemA" = itemA,
                       "similarity_itemA_causal_distance" = mean_causal_distance,
                       "similarity_itemA_freesort_distance" = mean_freesort_distance, 
                       "similar_itemA_freesort_minus_causal" = freesort_minus_causal),
              by = c("itemB", "similarity_itemA"))

  
trait_inference_condition <- d_model %>% 
  # select(-causal_itemA) %>% 
  filter(condition == "inference") %>% 
    left_join(d_freesort_and_causal_distances %>% 
                select(itemA, itemB, mean_causal_distance, mean_freesort_distance, freesort_minus_causal) %>% 
                rename("causal_itemA" = itemA,
                       "causal_itemA_causal_distance" = mean_causal_distance,
                       "causal_itemA_freesort_distance" = mean_freesort_distance,
                       "causal_itemA_freesort_minus_causal" = freesort_minus_causal),
              by = c("itemB", "causal_itemA")) %>%
    left_join(d_freesort_and_causal_distances %>% 
                select(itemA, itemB, mean_causal_distance, mean_freesort_distance, freesort_minus_causal) %>% 
                rename("similarity_itemA" = itemA,
                       "similarity_itemA_causal_distance" = mean_causal_distance,
                       "similarity_itemA_freesort_distance" = mean_freesort_distance, 
                       "similar_itemA_freesort_minus_causal" = freesort_minus_causal),
              by = c("itemB", "similarity_itemA"))

```

## Effect of condition

### run model

```{r}
#model
model_effect_of_condition <- glmer(responses ~ condition + (1|subject_id), data = d_model, family = binomial(link = 'logit'))
summary(model_effect_of_condition)
plot(allEffects(model_effect_of_condition))
#check_model(model_effect_of_condition) #binned residual issue?

#probabilities
effects_data <- allEffects(model_effect_of_condition)
effects_df <- as.data.frame(effects_data$condition)
effects_df

```

### predictions

model predictions for the probability of specific items, for writeup

```{r}
#see something
model_with_itemB <- glmer(responses ~ condition * (itemB == "see something") + (1|subject_id), 
                          data = d_model, family = binomial(link = 'logit'))
summary(model_with_itemB)

new_data <- data.frame(
  condition = c("inference", "intervention"),
  itemB = "see something",
  subject_id = d_model$subject_id[1]  
)

new_data$predicted_response <- predict(model_with_itemB, newdata = new_data, type = "response")
new_data


#think about something
model_with_itemB <- glmer(responses ~ condition * (itemB == "think about something") + (1|subject_id), 
                          data = d_model, family = binomial(link = 'logit'))
summary(model_with_itemB)

new_data <- data.frame(
  condition = c("inference", "intervention"),
  itemB = "think about something",
  subject_id = d_model$subject_id[1] 
)

new_data$predicted_response <- predict(model_with_itemB, newdata = new_data, type = "response")
new_data



```


## Effect of distances: trait inference condition

### run model

```{r}
# Modelling predictors jointly
#model_effect_of_distances_inference <- glmer(responses ~ causal_itemA_freesort_minus_causal + similar_itemA_freesort_minus_causal + (1|subject_id), data = trait_inference_condition, family = binomial(link = "logit")) # TODO figure out failure to converge here --> removed subj_id random effect

model_effect_of_distances_inference <- glm(responses ~ causal_itemA_freesort_minus_causal + similar_itemA_freesort_minus_causal, data = trait_inference_condition, family = binomial(link = "logit"))

summary(model_effect_of_distances_inference)
plot(allEffects(model_effect_of_distances_inference))
#check_model(model_effect_of_distances_inference)

```

### figure for paper

```{r}
effects_data <- allEffects(model_effect_of_distances_inference)

causal_effect <- as.data.frame(effects_data$`causal_itemA_freesort_minus_causal`)
similar_effect <- as.data.frame(effects_data$`similar_itemA_freesort_minus_causal`)

causal_effect$Predictor <- "Causal Item"
similar_effect$Predictor <- "Similar Item"

colnames(causal_effect)[1] <- "PredictorValue"
colnames(similar_effect)[1] <- "PredictorValue"

combined_effects <- rbind(causal_effect, similar_effect)


base_plot <- ggplot(combined_effects, aes(x = PredictorValue, y = fit, color = Predictor)) +
  geom_line(size = 1.2) +  # Adjust line size for emphasis
  geom_hline(yintercept = .5, linetype = "dashed") +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = Predictor), alpha = 0.2) +
  scale_color_manual(
    values = c("Causal Item" = "black", "Similar Item" = "grey")
  ) +
  scale_fill_manual(
    values = c("Causal Item" = "black", "Similar Item" = "grey")
  ) +
  labs(
    title = "Probability of Making A Causal Choice in Trait Inference Condition",
    x = "Distance between Choice and Target Item",
    y = "Probability",
    color = "Choice Item",
    fill = "Choice Item"
  ) +
  theme_cowplot() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 8),  
    axis.text.x = element_text(size = 20),  
    axis.text.y = element_text(size = 20),  
    axis.title.y = element_text(size = 20),  
    axis.title.x = element_text(size = 15),  
    legend.title = element_text(size = 20), 
    legend.text = element_text(size = 20),
    legend.position = c(0.5, 0.7) ,  #move legend inside the plot
    legend.background = element_rect(fill = "white", color = "black")  #border around the legend
  )

base_plot


```


## Effect of distances: intervention condition

### run model

```{r}
#Modelling both predictors 
model_effect_of_distances_intervention <- glmer(responses ~ causal_itemA_freesort_minus_causal + similar_itemA_freesort_minus_causal + (1|subject_id), data = intervention_condition, family = binomial(link = "logit"))
summary(model_effect_of_distances_intervention)
plot(allEffects(model_effect_of_distances_intervention))
#check_model(model_effect_of_distances_intervention )

fixed_effects <- fixef(model_effect_of_distances_intervention) #log odds
odds_ratios <- exp(fixed_effects)  #convert log-odds to odds ratios
probabilities <- 1 / (1 + exp(-fixed_effects)) #convert log odds to probability

#nb: fixed effects are intercept and slope. slope is change log odds of outcome for a unit increase in predictor
```

### figure for paper

```{r}
effects_data <- allEffects(model_effect_of_distances_intervention)

causal_effect <- as.data.frame(effects_data$`causal_itemA_freesort_minus_causal`)
similar_effect <- as.data.frame(effects_data$`similar_itemA_freesort_minus_causal`)

causal_effect$Predictor <- "Causal Item"
similar_effect$Predictor <- "Similar Item"

colnames(causal_effect)[1] <- "PredictorValue"
colnames(similar_effect)[1] <- "PredictorValue"

combined_effects <- rbind(causal_effect, similar_effect)

base_plot <- ggplot(combined_effects, aes(x = PredictorValue, y = fit, color = Predictor)) +
  geom_line(size = 1.2) +  # Adjust line size for emphasis
  geom_hline(yintercept = .5, linetype = "dashed") +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = Predictor), alpha = 0.2) +
  scale_color_manual(
    values = c("Causal Item" = "black", "Similar Item" = "grey")
  ) +
  scale_fill_manual(
    values = c("Causal Item" = "black", "Similar Item" = "grey")
  ) +
  labs(
    title = "Probability of Making A Causal Choice in Intervention Condition",
    x = "Distance between Choice and Target Item",
    y = "Probability",
    color = "Choice Item",
    fill = "Choice Item"
  ) +
  theme_cowplot() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 8),  
    axis.text.x = element_text(size = 20),  
    axis.text.y = element_text(size = 20),  
    axis.title.y = element_text(size = 20),  
    axis.title.x = element_text(size = 15),  
    legend.title = element_text(size = 20), 
    legend.text = element_text(size = 20),
    legend.position = c(0.5, 0.7) ,  #move legend inside the plot
    legend.background = element_rect(fill = "white", color = "black")  #border around the legend

    #plot.margin = margin(20, 20, 50, 20)
  )
base_plot

```
```{r}

```





